Intentional release of native species undermines ecological stability

### 2022-09-27 rmd/supporting_info.Rmd
Update supporting_info.Rmd

## ssm estimate
df_bp <- list.files(path = here::here("data_fmt"),
                    full.names = T,
                    pattern = "joint") %>% 
  readRDS() %>% 
  filter(param == "bp_value") %>% 
  dplyr::select(mean)

@@

df_mcmc <- lapply(c("summary_joint", "summary_reg", "summary_multi_ricker"),
       function(x) {
         list.files(path = here::here("output"),
                    full.names = T,
                    pattern = x) %>% 
           readRDS() %>% 
           ungroup() %>% 
           dplyr::select(n_total_mcmc,
                         n_sample,
                         n_thin,
                         n_burn) %>% 
           distinct() %>% 
           mutate(model = str_extract(x, "joint|reg|ricker"))
       }) %>% 
  bind_rows()

@@

We subsequently analyzed a whole community model with 10 species using important parameter combinations identified in the two-species simulation: intrinsic growth of enhanced species ($r_1 =$ `r unique(list_param[[2]]$r1)`), average strength of interspecific competition ($\bar{\alpha}=$ `r unique(list_param[[2]]$alpha)`), carrying capacity ($K=$ `r unique(list_param[[2]]$k)`). We confirmed that these parameters were influential in the ten-species community model via sensitivity analysis (**Table S2 -- S4**). We also varied relative fitness of captive-bred individuals ($f_R=$ `r unique(list_param[[2]]$phi)`) as this is a common interest in captive-breeding programs. Meanwhile, we fixed values of the following parameters: the number of species ($S=$ `r unique(list_param[[2]]$n_species)`), maximum intrinsic growth rate of unenhanced species ($r_{max}=$ `r unique(list_param[[2]]$r_max)`), and environmental stochasticity ($\sigma_{\varepsilon}=$ `r unique(list_param[[2]]$sd_env)`). This simulation setup resulted in `r nrow(list_param[[2]])` sets of parameter combinations.

@@

```{=tex}
\begin{align}
\begin{split}
&\ln n_{s,t} = \xi_{1,s} + \xi_{2,s} \ln n'_{s,t-1} + \varepsilon_{s,t}^{\text{state}}\\
&\ln n'_{s,t-1} = \frac{\sum_{q=1}^Q (\tau^{q} \ln n_{s,t-q})}{\sum_{q=1}^Q \tau^{q}}
\end{split}
(\#eq:state)
\end{align}
```
where $\xi_{1,s}$ is the site-specific growth rate at site $s$, $\xi_{2,s}$ the parameter characterizing density-dependence, and $\varepsilon_{s,t}^{\text{state}}$ the process error that follows a normal distribution as $\varepsilon_{s,t}^{\text{state}} \sim \text{Normal}(0, \sigma_{\text{state},s}^2)$. The parameters $\xi_{1,s}$ and $\xi_{2,s}$ were drawn from a multi-variate normal distribution as $\pmb{\xi} \sim \text{MVN}(\pmb{\theta_{\xi}}, \pmb{\Omega_{\xi}})$, where $\pmb{\xi}$ is the matrix of $\xi_{1,s}$ and $\xi_{2,s}$. Equation \@ref(eq:state) is structurally equivalent to a $Q$-order autoregressive (AR) model; however, our model has only one AR parameter $\tau$ (range: 0 -- 1) unlike ordinary AR models that require $Q$ AR parameters. This reduction of parameter numbers is possible because we express the AR parameters as a geometric-series $\tau^q$. Thus, the parameter $\tau$ controls how fast the density influence attenuates towards the past. In this study, we set $Q=3$ given the life span of the study species (typically 1 -- 3 years). We used median estimates of fish density $n_{s,t}$ to calculate the temporal CV, mean ($\mu$), and SD ($\sigma$) for each site. For a whole community, we summed the densities of enhanced and unenhanced species. We summarized the reconstructed community dynamics in **Figures S7 -- S9**.

We assessed the predictive performance of our model using the Bayesian p-value [@keryIntroductionWinBUGSEcologists2010], a value of which takes a range of 0 --1 and indicates over- (\~0.00), under- (\~1.00), or suitable-fitting (\~0.50) to the data. A Bayesian p-value for our state-space models was `r op(min(df_bp$mean))`, indicating that our model specification is appropriate.

@@

**Priors.** Vague priors were assigned to most parameters: $\text{Normal(0, 10)}$ for regression coefficients ($\gamma$ and $\gamma'$) and means ($\theta_{\beta}$ and $\pmb{\theta_{\xi}}$), $\text{Half-student's}~t(0, 2.5, 6)$ for SDs ($\sigma_{\beta}$, $\sigma_{\text{obs},s}$, $\sigma_{\text{state},s}$, $\sigma_{\text{obs},i}$, $\sigma_{\text{state},i}$ $\sigma_{\lambda}$, $\sigma_{\gamma}$, and $\sigma_{\epsilon,i}$), $\text{scaled-Inv-Wishart}(\pmb{\phi})$ ($\pmb{\phi}$ is a vector of prior scales $\pmb{\phi}=\{2.5,...,2.5\}$) with the degrees of freedom = 2 for variance-covariance matrices ($\pmb{\Omega_{\xi}}$, $\pmb{\Omega_y}$, and $\pmb{\Omega_{\varepsilon}}$), and $\text{Normal}(0, 1)$ for latent factors ($\zeta_{t,d}$). We used multiplicative gamma priors with parameters $\nu = 3$, $a_1=2$, and $a_2=3$ (see Durante [@duranteNoteMultiplicativeGamma2017] for interpretation) for latent factors ($\delta_{d,i}$) following Ovaskainen *et al.* [@ovaskainenHowAreSpecies2017]. We assigned a weakly informative priors to initial fish densities $\text{ln}~n_{i,t}$ ($t = 1, 2, 3$) using $\text{Student's}~t(-2, 2, 6)$ to enhance model convergence.

**Model fitting.** We fitted the models to the data using JAGS version 4.1.0 through *runjags* package version 2.2.0-2 in R [@denwoodRunjagsPackageProviding2016]. We assigned weakly informative priors to parameters (**Table S6**). Four Markov chain Monte Carlo (MCMC) chains were run until parameter estimates converged. The total MCMC iteration was XXX. Convergence was assessed by examining whether the $\hat{R}$ indicator of each parameter approached \< 1.1 [@gelmanDataAnalysisUsing2007]. Data manipulation and analysis were performed in R version 4.2.1 [@rcoreteamLanguageEnvironmentStatistical2021]. Parameter estimates were summarized in **Table S7 -- S10**.

@@

# Figure S1-6
## load figures
source(here::here("code/si_figure_theory.R"))

@@

# Figure S7-9
## load figures
source(here::here("code/si_figure_obs_dyns.R"))
j <- 6

@@

## Figure S10 Environmental variables

```{r figure_s10, fig.width=11, fig.height=8}

@@

## Figure S11 Correlation plot

```{r figure_s11, fig.width=11, fig.height=11}

>>>>

## ssm estimate
df_bp <- list.files(path = here::here("output"),
                    full.names = T,
                    pattern = "summary_ssm_ar") %>% 
  readRDS() %>% 
  filter(param == "bp_value") %>% 
  dplyr::select(mean)

@@

df_mcmc <- lapply(c("summary_ssm_ar", "summary_reg", "summary_multi_ricker"),
       function(x) {
         list.files(path = here::here("output"),
                    full.names = T,
                    pattern = x) %>% 
           readRDS() %>% 
           ungroup() %>% 
           dplyr::select(n_total_mcmc,
                         n_sample,
                         n_thin,
                         n_burn) %>% 
           distinct() %>% 
           mutate(model = str_extract(x, "ssm|reg|ricker"))
       }) %>% 
  bind_rows()

@@

We subsequently analyzed a whole community model with 10 species using important parameter combinations identified in the two-species simulation: intrinsic growth of enhanced species ($r_1 =$ `r unique(list_param[[2]]$r1)`), average strength of interspecific competition ($\bar{\alpha}=$ `r unique(list_param[[2]]$alpha)`), carrying capacity ($K=$ `r unique(list_param[[2]]$k)`). We also varied relative fitness of captive-bred individuals ($f_R=$ `r unique(list_param[[2]]$phi)`) as this is a common interest in captive-breeding programs. Meanwhile, we fixed values of the following parameters: the number of species ($S=$ `r unique(list_param[[2]]$n_species)`), maximum intrinsic growth rate of unenhanced species ($r_{max}=$ `r unique(list_param[[2]]$r_max)`), and environmental stochasticity ($\sigma_{\varepsilon}=$ `r unique(list_param[[2]]$sd_env)`). This simulation setup resulted in `r nrow(list_param[[2]])` sets of parameter combinations.

@@

```{=tex}
\begin{align}
\begin{split}
&\ln n_{s,t} = \xi_{1,s} + \xi_{2,s} \ln n'_{s,t-1} + \varepsilon_{s,t}^{\text{state}}\\
&\ln n'_{s,t-1} = \frac{\sum_{q=1}^Q (\tau^{q} \ln n_{s,t-q})}{\sum_{q=1}^Q \tau^{q}}
\end{split}
(\#eq:state)
\end{align}
```
where $\xi_{1,s}$ is the site-specific constant at site $s$, $\xi_{2,s}$ the parameter characterizing an autoregressive (AR) process, and $\varepsilon_{s,t}^{\text{state}}$ the process error that follows a normal distribution as $\varepsilon_{s,t}^{\text{state}} \sim \text{Normal}(0, \sigma_{\text{state},s}^2)$. The parameters $\xi_{1,s}$ and $\xi_{2,s}$ were drawn from a multi-variate normal distribution as $\pmb{\xi} \sim \text{MVN}(\pmb{\theta_{\xi}}, \pmb{\Omega_{\xi}})$, where $\pmb{\xi}$ is the matrix of $\xi_{1,s}$ and $\xi_{2,s}$. Equation \@ref(eq:state) is structurally equivalent to a $Q$-order AR model; however, we expressed the part of AR parameters as a geometric-series $\tau^q$ to reduce the number of parameters (our model has two AR parameters $\xi_{2,s}$ and $\tau$ regardless of order $Q$). The parameter $\tau$ controls how fast the density influence attenuates towards the past [@takagiTimescaleDependencyHost2015]. In this study, we set $Q=3$ given the life span of the study species (typically 1 -- 3 years). We used median estimates of fish density $n_{s,t}$ to calculate the temporal CV, mean ($\mu$), and SD ($\sigma$) for each site. For a whole community, we summed the densities of enhanced and unenhanced species. We summarized the reconstructed community dynamics in **Figures S7 -- S9**.

We assessed the predictive performance of our model using the Bayesian p-value [@keryIntroductionWinBUGSEcologists2010], a value of which takes a range of 0 --1 and indicates over- (\~0.00), under- (\~1.00), or suitable-fitting (\~0.50) to the data. A Bayesian p-value for our state-space models was `r op(round(df_bp$mean, 2))`, indicating that our model specification is appropriate.

@@

**Model fitting.** We fitted the models to the data using JAGS version 4.1.0 through *runjags* package version 2.2.0-2 in R [@denwoodRunjagsPackageProviding2016]. We assigned vague or weakly informative priors to parameters (**Table S3**). Four Markov chain Monte Carlo (MCMC) chains were run until parameter estimates converged. The total MCMC iterations ranged from `r min(df_mcmc$n_total_mcmc)` to `r max(df_mcmc$n_total_mcmc)`, in which MCMC samples were saved every `r unique(df_mcmc$n_thin)` steps for the calculation of posterior distributions after the initial `r max(df_mcmc$n_burn)` burn-in period. Convergence was assessed by examining whether the $\hat{R}$ indicator of each parameter approached $<1.1$ [@gelmanDataAnalysisUsing2007]. Data manipulation and analysis were performed in R version 4.2.1 [@rcoreteamLanguageEnvironmentStatistical2021]. Parameter estimates were summarized in **Table S7 -- S10**.

@@

# Figure S1-8
## load figures
source(here::here("code/si_figure_theory.R"))

@@

# Figure S9-11
## load figures
source(here::here("code/si_figure_obs_dyns.R"))
j <- 8

@@

## Figure S12 Environmental variables

```{r figure_s12, fig.width=11, fig.height=8}

@@

## Figure S13 Correlation plot

```{r figure_s13, fig.width=11, fig.height=11}
